\bf \texttt{forestecology} package for modeling interspecies competition between trees

### 2021-01-13 etc/paper/forestecology.tex
Reknit manuscript

  \begin{itemize}
  \tightlist
  \item
    fitting models for interspecific competition and assessing them
    using spatial crossvalidation
  \item
    Leverage ForestGEO protocols providing standardization
  \end{itemize}
\item
  Indicate the approach and the methods: Use tidyverse, simple features,
  and tidymodels packages.

  \begin{itemize}
  \tightlist
  \item
    tidyverse: as stated in tidy tools manifesto: standadized data
    structures, functional programming with pipe, designed for humans
  \item
    sf: tidyverse-friendly package that makes wrangling and visualizing
    spatial data much easier
  \item
    tidymodels: given our spatial-crossvalidation, use tidymodels
    framework is a collection of packages for modeling and machine
    learning using tidyverse principles
  \end{itemize}
\item
  Outline the main results: We replicate from scratch the figure in
  PLOSOne paper using Big Woods data, conduct similar analysis for SCBI
  data.

@@

  \begin{itemize}
  \tightlist
  \item
    New scientific conclusions from SCBI data
  \item
    this can serve as blue print for other modeling situations.
  \item
    Modularly switch out our bayesian lm() functions to anything you
    want.
  \end{itemize}
\end{enumerate}

@@

Repeat-censused forest plots offer excellent data to test neighborhood
models of tree competition \citet{allen_permutation_2020}
\citet{canham_neighborhood_2006} \citet{uriarte_spatially_2004}. Here we
describe an R package, \texttt{forestecology}, to do that. This package
implements the methods in \citet{allen_permutation_2020}. It provides: a
convenient way specify and fit models of tree growth based on
neighborhood competition; a spatial cross validation method to test and
compare model fits \citet{roberts_cross-validation_2017}; and an
ANOVA-like method to assess whether the competitor identity matters in
these models. The model is written to work with ForestGEO plot data
\citet{andersonteixeira_ctfs-forestgeo_2015}, but we envision that it
could easily be modified to work with data from other forest plots,
e.g.~the US Forest Service Forest Inventory and Analysis plots
\citet{smith_forest_2002}.

@@

We demonstrate the \texttt{forestecology} package's features on two data
sets, both based on inventory censuses of two sites from the Smithsonian
Institution's ForestGEO international network of 72 long‐term forest
dynamics research sites \citet{andersonteixeira_ctfs-forestgeo_2015}.
First, the Michigan Big Woods Forest Dynamics Plot located at the Edwin
S. George Reserve in Pinckney, MI, USA. The 23 ha plot is situated in
mature oak-hickory forest. The canopy is dominated by white oak (Quercus
alba), northern red oak (Quercus rubra), black oak (Quercus velutina),
shagbark hickory (Carya ovata) and pignut hickory (Carya glabra). The
most common understory tree is witch-hazel (Hamamelis virginiana)
\citet{allen_michigan_2020}. In the example below, we will preface any
data frames from this plot in with \texttt{bw\_}.

Second, the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot, located at the Smithsonian's National Zoo and
Conservation Biology Institute in Front Royal, VA, USA. The 25.6 ha (640
x 400 m) plot is located at the intersection of three of the major
physiographic provinces of the eastern US: the Blue Ridge, Ridge and
Valley, and Piedmont provinces and is adjacent to the northern end of
Shenandoah National Park. The forest type is typical mature secondary
eastern mixed deciduous forest, with a canopy dominated by tulip poplar
(Liriodendron tulipifera), oaks (Quercus spp.), and hickories (Carya
spp.), and an understory composed mainly of spicebush (Lindera benzoin),
paw-paw (Asimina triloba), American hornbeam (Carpinus caroliniana), and
witch hazel (Hamamelis virginiana) \citet{bourg_initial_2013}. In the
example below, we will preface any data frames from this plot in with
\texttt{scbi\_}.

@@

In the case of Big Woods data, we first remove all trees that were
re-sprouts in the later (2014) census. Additionally, we have included
three classification of tree species: \texttt{species}, \texttt{family},
and \texttt{trait\_group}. DESCRIBE THESE

@@

We now encode spatial information to the \texttt{growth\_df} data
frames. First, in order to control for study region edge effects, we add
``buffers'' to the periphery of the study region (cite Waller?). Our
model of interspecific competition relies on a spatial definition of who
the competitor trees are for focal trees of interest. Since certain
explanatory variables such as biomass are cumulative, we must ensure
that all trees being modeled are not biased to have different neighbor
structures. This is a particular concern for trees at the boundary of
study regions, which will not have the same number of neighbors as trees
in the internal part of the study region.

Second, our ultimate method for model assessment will rely on estimates
of model error as generated by cross-validation. Conventional
cross-validation schemes assign observations to folds by resampling
individual observations at random. However, underlying this scheme is an
assumption that the observations are independent. In the case of forest
census data, observations exhibit spatial autocorrelation, and thus this
dependence must be incorporated in our resampling scheme in spatial
cross-validation \citet{roberts2017} \citet{pohjankukka2017}. We will
therefore associate portions of the study region to spatial folds.

@@

The second constant is \texttt{cv\_fold\_size} which defines the length
and width of the spatial folds (note that for now the spatial folds are
restricted be squares). We will then use this constant to associate each
observed tree to one of \(k\) folds in the respective study region. In
the Big Woods example below we will use the \texttt{blockCV} R package
that has implemented spatial cross-validation while for the SCBI we will
do this manually \citet{valavi2019}.

>>>>

  \begin{itemize}
  \tightlist
  \item
    (Eventually) modularly fitting models for interspecific competition
    and assessing them using spatial crossvalidation
  \item
    Leverage ForestGEO protocols providing standardization
  \end{itemize}
\item
  Indicate the approach and the methods: Use tidyverse, simple features,
  and tidymodels packages.

  \begin{itemize}
  \tightlist
  \item
    tidyverse: as stated in tidy tools manifesto: standadized data
    structures, functional programming with pipe, designed for humans
  \item
    sf: tidyverse-friendly package that makes wrangling and visualizing
    spatial data much easier
  \item
    tidymodels: given our spatial-crossvalidation, use tidymodels
    framework is a collection of packages for modeling and machine
    learning using tidyverse principles
  \item
    Dave: hmm, I think here we might need to mention the methods we
    implement. Something like, the package provides functions to
    specific a linear, bayesian neighborhood competition model of
    growth, then fit the model, and compare competing models with
    spatiall cross validation. Or somethign like that.
  \end{itemize}
\item
  Outline the main results: We replicate from scratch the figure in
  PLOSOne paper using Big Woods data, conduct similar analysis for SCBI
  data.

@@

  \begin{itemize}
  \tightlist
  \item
    New scientific conclusions from SCBI data
  \item
    Modularly switch out our bayesian lm() functions to anything you
    want.
  \item
    this can serve as blue print for other modeling situations.
  \end{itemize}
\end{enumerate}

@@

Repeat-censused forest plots offer excellent data to test neighborhood
models of tree competition \citet{allen_permutation_2020}
\citet{canham_neighborhood_2006} \citet{uriarte_spatially_2004}. Here we
describe an R package, \texttt{forestecology}, to do that. This package
implements the methods in \citet{allen_permutation_2020}. It provides: a
convenient way to specify and fit models of tree growth based on
neighborhood competition; a spatial cross validation method to test and
compare model fits \citet{roberts_cross-validation_2017}; and an
ANOVA-like method to assess whether the competitor identity matters in
these models. The model is written to work with ForestGEO plot data
\citet{andersonteixeira_ctfs-forestgeo_2015}, but we envision that it
could easily be modified to work with data from other forest plots,
e.g.~the US Forest Service Forest Inventory and Analysis plots
\citet{smith_forest_2002}.

@@

We demonstrate the \texttt{forestecology} package's features on two data
sets, both based on inventory censuses of two sites from the Smithsonian
Institution's ForestGEO international network of 72 long‐term forest
dynamics research sites \citet{andersonteixeira_ctfs-forestgeo_2015}.
First, the Michigan Big Woods Forest Dynamics Plot located at the Edwin
S. George Reserve in Pinckney, MI, USA. The 23 ha plot is situated in
mature oak-hickory forest. The canopy is dominated by white oak
(\emph{Quercus alba}), northern red oak (\emph{Q. rubra}), black oak
(\emph{Q. velutina}), shagbark hickory (\emph{Carya ovata}) and pignut
hickory (\emph{C. glabra}) \citet{allen_michigan_2020}. In the example
below, we will preface any data frames from this plot in with
\texttt{bw\_}.

Second, the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot, located at the Smithsonian's National Zoo and
Conservation Biology Institute in Front Royal, VA, USA. The 25.6 ha (640
x 400 m) plot is located at the intersection of three of the major
physiographic provinces of the eastern US: the Blue Ridge, Ridge and
Valley, and Piedmont provinces and is adjacent to the northern end of
Shenandoah National Park. The forest type is typical mature secondary
eastern mixed deciduous forest, with a canopy dominated by tulip poplar
(\emph{Liriodendron tulipifera}), oaks (\emph{Quercus} spp.), and
hickories (\emph{Carya} spp.), and an understory composed mainly of
spicebush (\emph{Lindera benzoin}), paw-paw (\emph{Asimina triloba}),
American hornbeam (\emph{Carpinus caroliniana}), and witch hazel
(\emph{Hamamelis virginiana}) \citet{bourg_initial_2013}. In the example
below, we will preface any data frames from this plot in with
\texttt{scbi\_}.

@@

In the case of Big Woods data, we first remove all trees that were
re-sprouts in the later (2014) census. Additionally, we have included
two classification of tree species: \texttt{species} and
\texttt{family}. To illustrate model comparison, we test competition
models in which individuals are grouped by species and by family.

@@

We now encode spatial information to the \texttt{growth\_df} data
frames. First, in order to control for study region edge effects, we add
``buffers'' to the periphery of the study region (cite Waller?). Our
model of interspecific competition relies on a spatial definition of who
the competitor trees are for focal trees of interest. Since certain
explanatory variables such as basal area are cumulative, we must ensure
that all trees being modeled are not biased to have different neighbor
structures. This is a particular concern for trees at the boundary of
study regions, which will not have the same number of neighbors as trees
in the internal part of the study region.

Second, our ultimate method for model assessment will rely on estimates
of model error as generated by cross-validation. Conventional
cross-validation schemes assign observations to folds by resampling
individual observations at random. However, underlying this scheme is an
assumption that the observations are independent. In the case of forest
census data, observations exhibit spatial autocorrelation, and thus this
dependence must be incorporated in our resampling scheme in spatial
cross-validation \citet{roberts_cross-validation_2017}
\citet{pohjankukka_estimating_2017} We will therefore associate portions
of the study region to spatial folds.

@@

The second constant is \texttt{cv\_fold\_size} which defines the length
and width of the spatial folds (note that for now the spatial folds are
restricted be squares). We will then use this constant to associate each
observed tree to one of \(k\) folds in the respective study region. In
the Big Woods example below we will use the \texttt{blockCV} R package
that has implemented spatial cross-validation while for the SCBI we will
do this manually \citet{valavi_blockcv_2019}
